---
title: "Models of Neighborhood Change in San Diego"
author: "Melanie Phoenix"
date: "2022-11-30"
output: 
  flexdashboard::flex_dashboard:
    theme: readable
    source: embed
    smart: false 
runtime: shiny
---

```{r global, echo=FALSE}
# LOAD PACKAGES

library(cartogram)     #spatial maps w/ tract size bias reduction
library(corrplot)
library(datasets)
library(devtools)
library(dplyr)         #data wrangling
library(DT)            #formatting output
library(flexdashboard) #dashboard layout, widgets, and uploading
library(geojsonio)     #maps
library(ggmap)         #maps
library(ggplot2)       #graphing 
library(ggthemes)      #nice formats for ggplots
library(here)
library(import)
library(knitr)         #formatting output
library(leaflet)       #maps
library(magrittr)
library(maptools)      #spatial object manipulation
library(mclust)        #cluster analysis 
library(pals)          #maps
library(pander)        #formatting output
library(rgdal)         #maps
library(rgeos)         #Interface to open-source geometry engine (GEOS)
library(rsconnect)     #dashboard layout, widgets, and uploading
library(scales)
library(sf)            #maps
library(shiny)         #dashboard layout, widgets, and uploading
library(sp)            #maps
library(stargazer)     #formatting output
library(tidycensus)
library(tidyverse)
library(tigris)
library(tmap)          #maps
library(viridis)       #maps
```


```{r, warning=FALSE, include=FALSE}
# DATA STEPS

# from local file path
san <- geojson_read( "san.geojson", what="sp" )

san <- spTransform( san, CRS("+init=epsg:3395") )

bb <- st_bbox( c( xmin = -13055498, xmax = -13011238, 
                  ymax = 3853073, ymin = 3817112 ), 
               crs = st_crs("+init=epsg:3395"))
# reproject the map 
san2 <- spTransform( san, CRS("+init=epsg:3395") )

# convert the sp map format to 
# an sf (simple features) format:
# ggmap requires the sf format
san.sf <- st_as_sf( san2 )

# separate out the data frame from the map
sdd <- as.data.frame( san.sf )

URL1 <- "https://github.com/DS4PS/cpp-529-fall-2020/raw/main/LABS/data/rodeo/LTDB-2000.rds"
d5 <- readRDS( gzcon( url( URL1 ) ) )

URL2 <- "https://github.com/DS4PS/cpp-529-fall-2020/raw/main/LABS/data/rodeo/LTDB-2010.rds"
d6 <- readRDS( gzcon( url( URL2 ) ) )

URLmd <- "https://github.com/DS4PS/cpp-529-fall-2020/raw/main/LABS/data/rodeo/LTDB-META-DATA.rds"
md <- readRDS( gzcon( url( URLmd ) ) )

d5 <- select( d5, - year )
d6 <- select( d6, - year )

d <- merge( d5, d6, by="tractid" )
d <- merge( d, md, by="tractid" )

table( d$urban )

d <- filter( d, urban == "urban" )

d <- select( d, tractid, 
             mhmval00, mhmval12, 
             hinc00, 
             hu00, vac00, own00, rent00, h30old00,
             empclf00, clf00, unemp00, prof00,  
             dpov00, npov00,
             ag25up00, hs00, col00, 
             pop00.x, nhwht00, nhblk00, hisp00, asian00,
             cbsa, cbsaname )

 
d <- 
  d %>%
  mutate( # percent white in 2000
          p.white = 100 * nhwht00 / pop00.x,
          # percent black in 2000
          p.black = 100 * nhblk00 / pop00.x,
          # percent hispanic in 2000
          p.hisp = 100 * hisp00 / pop00.x, 
          # percent asian in 2000
          p.asian = 100 * asian00 / pop00.x,
          # percent high school grads by age 25 in 2000 
          p.hs = 100 * (hs00+col00) / ag25up00,
          # percent pop with college degree in 2000
          p.col = 100 * col00 / ag25up00,
          # percent employed in professional fields in 2000
          p.prof = 100 * prof00 / empclf00,
          # percent unemployment  in 2000
          p.unemp = 100 * unemp00 / clf00,
          # percent of housing lots in tract that are vacant in 2000
          p.vacant = 100 * vac00 / hu00,
          # dollar change in median home value 2000 to 2010 
          pov.rate = 100 * npov00 / dpov00)
          


# adjust 2000 home values for inflation 
mhv.00 <- d$mhmval00 * 1.28855  
mhv.10 <- d$mhmval12

# change in MHV in dollars
mhv.change <- mhv.10 - mhv.00


# drop low 2000 median home values
# to avoid unrealistic growth rates.
#
# tracts with homes that cost less than
# $1,000 are outliers
mhv.00[ mhv.00 < 1000 ] <- NA

# change in MHV in percent
mhv.growth <- 100 * ( mhv.change / mhv.00 )

d$mhv.00 <- mhv.00
d$mhv.10 <- mhv.10
d$mhv.change <- mhv.change
d$mhv.growth <- mhv.growth 

# geoid-01 is the hypothetical name of tract ID in the shapefile
# geoid-02 is the hypothetical name of tract ID in the census dataset

# dorling must be an sp shapefile 

```


Community Demographics  
=====================================  

Inputs {.sidebar}
-------------------------------------

```{r}

these.variables <- c("pnhwht12", "pnhblk12", "phisp12", "pntv12", "pfb12", "polang12", 
"phs12", "pcol12", "punemp12", "pflabf12", "pprof12", "pmanuf12", 
"pvet12", "psemp12", "hinc12", "incpc12", "ppov12", "pown12", 
"pvac12", "pmulti12", "mrent12", "mhmval12", "p30old12", "p10yrs12", 
"p18und12", "p60up12", "p75up12", "pmar12", "pwds12", "pfhh12")

dd.URL <- "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/data-dictionary.csv"
data.dictionary <- read.csv( dd.URL, stringsAsFactors=F ) 
# data.dictionary must have "LABEL" for var name and "VARIABLE" for description
data.dictionary <- data.dictionary[ -1 , ] # drop the original column names
data.label <- data.dictionary[ data.dictionary$VARIABLE %in% these.variables , ]

# replace these with descriptive labels 
# from the data dictionary 
temp.names <- paste0( "Variable ", these.variables )
#
Short.names <- c("Caucasian %", "African American %", "Hispanic %", "Native American %", "Foreign-Born %", "Non-English Homes %", "High School %", "College Degree %", "Unemployed %", "Female Worker %", "Professional  %", "Manufacturer %", "Veteran %", "Entrepreneur %", "Med Home Income", "Per Capita Income", "Poverty %", "Home Owner %", "Vacant %", "Multi-Family Homes %", "Med Rent", "Med Home Value", "30+ yr old Homes %", "HH under 10yr Homes %", "Below 17yrs %", "60+ yrs %", "75+ yrs %", "Married %", "Separated %", "Female HH %")
# Adding interpretable variable names
# from the data dictionary:
# add a name attribute for each variable
# 
# value <- c(1,2,3)
# dd.name <- c("one","two","three")
value <- data.label$LABEL
dd.name <- data.label$VARIABLE
# x <- dd.name
# names(x) <- value
xnames <- dd.name
names(xnames) <- value
# dd names and values linked
# names( x[2] )
# can now get the label using the value
# using the name attributes 
# x[ "two" ]
#
# to add labels to the maps
# use the radio button value 
# to get the data dictionary label: 
# x[input$demographics ]
# x[input$label 
names(these.variables) <- c("Caucasian %", "African American %", "Hispanic %", "Native American %", "Foreign-Born %", "Non-English Homes %", "High School %", "College Degree %", "Unemployed %", "Female Worker %", "Professional  %", "Manufacturer %", "Veteran %", "Entrepreneur %", "Med Home Income", "Per Capita Income", "Poverty %", "Home Owner %", "Vacant %", "Multi-Family Homes %", "Med Rent", "Med Home Value", "30+ yr old Homes %", "HH under 10yr Homes %", "Below 17yrs %", "60+ yrs %", "75+ yrs %", "Married %", "Separated %", "Female HH %")
temp.names <- Short.names
#
radioButtons( inputId="Demographics", 
              label = h3("Census Variables"),
              choices =  these.variables, 
#              choiceNames= data.label$VARIABLE,
#              choiceValues= these.variables,
              selected="Unemployed %")
```

```



Row {.tabset}
-------------------------------------


### Choropleth Map San Diego


```{r} 

renderPlot({

# split the selected variable into deciles 
  
get_data <- 
  reactive({
             san.sf <- 
             san.sf %>% 
             mutate( q = ntile( get(input$Demographics), 10 ) )  
          })

  
ggplot( get_data() ) +
    geom_sf( aes( fill = q ), color=NA ) +
    coord_sf( datum=NA ) +
    labs( title = paste0( "Choropleth of Select Demographics: ", toupper(Short.names[which(input$demographics == these.variables)])),
          caption = "Source: Harmonized Census Files",
          fill = "Population Deciles" ) +
    scale_fill_gradientn( colours=rev(ocean.balance(10)), guide = "colourbar" ) + 
    xlim( xmin = -13055498, xmax = -13011238 ) + 
    ylim( ymin = 3817112, ymax = 3853073 )

})

```


### Variable Distribution 

```{r}

renderPlot({

# extract vector x from the data frame 
# x <-  d[ "punemp12" ] %>% unlist()

  
get_variable_x <- reactive({ sdd[ input$demographics ] })

x <- get_variable_x() %>% unlist()

cut.points <- quantile( x, seq( 0, 1, 0.1 ) )

hist( x, breaks=50, 
      col="gray", border="white", yaxt="n",
      main=paste0( "Histogram of variable ", toupper(Short.names[which( input$demographics == these.variables )])),
      xlab="red lines represent decile cut points" )

abline( v=cut.points, col="darkred", lty=3, lwd=2 )


})

```

Neighborhoods  
===================================== 

### Clusters  

```{r}

# define the bounding box corners 
bb <- st_bbox( c( xmin = -13055498, xmax = -13011238, 
                  ymax = 3853073, ymin = 3817112 ), 
               crs = st_crs("+init=epsg:3395"))


# ADD YOUR CUSTOM LABELS TO THE CLUSTERS 

san2$cluster[ san2$cluster == "1" ] <- "Native American Veterans"
san2$cluster[ san2$cluster == "2" ] <- "Girl Power"
san2$cluster[ san2$cluster == "3" ] <- "Black Girl Magic"
san2$cluster[ san2$cluster == "4" ] <- "Old School"
san2$cluster[ san2$cluster == "5" ] <- "Habla en Espano"
san2$cluster[ san2$cluster == "6" ] <- "Cali Life"


renderTmap({

tmap_mode("view")
tm_basemap( "CartoDB.Positron" )
tm_shape( san2, bbox=bb ) + 
  tm_polygons( col="cluster", palette="Accent",
               title="Community Types" ) 

})

```




NH Change 2000-2010  
===================================== 



Inputs {.sidebar}
-------------------------------------

```{r}

button.labels <- c("Median Home Value 2000","Median Home Value 2010","Value Change 2000-2010","Growth in Home Value")
button.values <- c("mhv.00", "mhv.10", "mhv.change", "mhv.growth")

  
radioButtons( inputId="home.value", 
              label = h3("Home Values"),
              choices = button.values, 
              choiceNames=button.labels,
              choiceValues=button.values,
              selected="Median Home Value 2000")

```



Row {.tabset}
-------------------------------------



### Home Values 

```{r}
# Median Home Values 2000
home.value <- function(MHV2000, MHV2010, CHV0010, GHV0010)
  
home.value %>% 
  {
  MHV2000 <- 
  {
hist( mhv.00, breaks=200, xlim=c(0,500000), 
      col="gray20", border="white",
      axes=F, 
      xlab="MHV (median = $138k)",
      ylab="",
      main="Median Home Value in 2000 (2010 US dollars)" )

axis( side=1, at=seq(0,500000,100000), 
      labels=c("$0","$100k","$200k","$300k","$400k","$500k") )

abline( v=median( mhv.00, na.rm=T ), col="orange", lwd=3 ) 
renderPlot({})
}
# split the selected variable into deciles 
#
#
# Median Home Values 2010
MHV2010 <- 
  {
hist( mhv.10, breaks=200, xlim=c(0,500000), 
      col="gray20", border="white",
      axes=F, 
      xlab="MHV (median = $138k)",
      ylab="",
      main="Median Home Value in 2010 (2010 US dollars)" )

axis( side=1, at=seq(0,500000,100000), 
      labels=c("$0","$100k","$200k","$300k","$400k","$500k") )

abline( v=median( mhv.10, na.rm=T ), col="orange", lwd=3 ) 
renderPlot({})
}
# 
#
# Change in home values 2000-2010
CHV0010 <-
  {
hist( mhv.change/1000, breaks=500, 
      xlim=c(-100,500), yaxt="n", xaxt="n",
      xlab="Thousand of US Dollars (adjusted to 2010)", cex.lab=1.5,
      ylab="", main="Change in Median Home Value 2000 to 2010",
      col="gray20", border="white" )

axis( side=1, at=seq( from=-100, to=500, by=100 ), 
      labels=paste0( "$", seq( from=-100, to=500, by=100 ), "k" ) )
        
mean.x <- mean( mhv.change/1000, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=200, y=1500, 
      labels=paste0( "Mean = ", dollar( round(1000*mean.x,0)) ), 
      col="darkorange", cex=1.2, pos=3 )

median.x <- median( mhv.change/1000, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=200, y=2000, 
      labels=paste0( "Median = ", dollar( round(1000*median.x,0)) ), 
      col="dodgerblue", cex=1.2, pos=3 )
}
# 
#
# Growth in Home Value
GHV0010 <-
  {
hist( mhv.growth, breaks=5000, 
      xlim=c(-100,200), yaxt="n", xaxt="n",
      xlab="", cex.main=1.5,
      ylab="", main="Growth in Home Value by Census Tract 2000 to 2010",
      col="gray40", border="white" )

axis( side=1, at=seq( from=-100, to=200, by=50 ), 
      labels=paste0( seq( from=-100, to=200, by=50 ), "%" ) )

ymax <- max(  )
        
mean.x <- mean( mhv.growth, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=100, y=(0.5*ymax), 
      labels=paste0( "Mean = ", round(mean.x,0), "%"), 
      col="darkorange", cex=1.2, pos=4 )

median.x <- median( mhv.growth, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=100, y=(0.6*ymax), 
      labels=paste0( "Median = ", round(median.x,0), "%"), 
      col="dodgerblue", cex=1.2, pos=4 )
  }
}

```


```{r}

renderPlot({

# split the selected variable into deciles 
  
get_data <- 
  reactive({
             san.sf <- 
             san.sf %>% 
             mutate( q = ntile( get(input$home.value), 10 ) )  
          })

  
ggplot( get_data() ) +
    geom_sf( aes( fill = q ), color=NA ) +
    coord_sf( datum=NA ) +
    labs( title = paste0( "Spatial Distribution of Home Values: ", toupper(Short.names[which( input$demographics == these.variables )])),
          caption = "Source: Harmonized Census Files",
          fill = "Home Value Deciles" ) +
    scale_fill_gradientn( colours=rev(ocean.balance(10)), guide = "colourbar" ) + 
    xlim( xmin = -13055498, xmax = -13011238 ) + 
    ylim( ymin = 3817112, ymax = 3853073 )

})

```


### Variable Distribution 

```{r}

renderPlot({

# extract vector x from the data frame 
# x <-  d[ "pnhwht12" ] %>% unlist()

  
get_variable_x <- reactive({ sdd[ input$home.value ] })

x <- get_variable_x() %>% unlist() %>% as.numeric()

cut.points <- quantile( x, seq( 0, 1, 0.1 ) )

hist( x, breaks=50, 
      col="gray", border="white", yaxt="n",
      main=paste0( "Histogram of ", toupper( input$home.value ) ),
      xlab="red lines represent decile cut points" )

abline( v=cut.points, col="darkred", lty=3, lwd=2 )


})

```  



Drivers of Change   
===================================== 



Inputs {.sidebar}
-------------------------------------

```{r}

Home_value <- c("mhv.2000","mhv.2010","mhv.change","mhv.growth")
HV_Label <- c("Median Home Value 2000","Median Home Value 2010","Value Change 2000-2010","Growth in Home Value")

button.labels <- HV_Label
button.values <- Home_value
  
radioButtons( inputId="dv", 
              label = h3("Select Your Dependent Variable"),
              choiceNames=button.labels,
              choiceValues=button.values,
              selected="mhv.change")


covariates <- c("pnhwht12", "pnhblk12", "phisp12", "pntv12", "pfb12", "polang12", 
"phs12", "pcol12", "punemp12", "pflabf12", "pprof12", "pmanuf12", 
"pvet12", "psemp12", "hinc12", "incpc12", "ppov12", "pown12", 
"pvac12", "pmulti12", "mrent12", "mhmval12", "p30old12", "p10yrs12", 
"p18und12", "p60up12", "p75up12", "pmar12", "pwds12", "pfhh12")

covariate.labels <- Short.names

checkboxGroupInput( inputId="covariates", 
              label = h3("Select Variables for Your Model"),
              choices = covariates,
              choiceNames=covariate.labels,
              choiceValues=covariates,
              selected=c("pnhwht12","pprof12","pvac12") )

```




Row {.tabset}
-------------------------------------



### Predicting Change 

```{r, results="asis"}

# RUNNING A REGRESSION WITH USER INPUTS
#
# create a formula object
# by constructing the formula from user selections 
# as a string then casting as a formula object 
# x.s <- c("x1","x2","x3" )
# formula.text <- paste( "y", " ~ ", paste0( x.s, collapse=" + ") )
# formula.object <- as.formula( formula.text )
#
# lm( formula.object, data=d )
#
# make sure all variables are in d


  
# check boxes return vectors 
get_covariates <-  reactive({ input$covariates })

DVhouses <- reactive({input$dv})

renderUI({
  
covariates <- get_covariates() 
formula.text <- paste0( DVhouses, " ~ ", paste( covariates, collapse=" + " ) )
fo <- as.formula( formula.text )
fo <- as.formula( formula.text )

m <- lm( fo, data=d )


# HTML( "<div style='width: 60%; margin: 0px auto;'>" )
HTML( 
  
  c("<br><br><br>",
  "<div type='regression' style='width: 60%; margin: 0px auto;'>",
  stargazer( m, type="html", omit.stat=c("rsq","f") ),
  "</div>",
  "<br><br><br>"
  )
  
  )


})

# HTML( reg.table )

```



### Correlation Plots 

```{r}

renderPlot({
  dcorr <- data.frame(covariates, DVhouses())
  pairs( dcorr )
})


```

<style>

.chart-shim { overflow: auto; }
 
table{
   border-spacing:1px;
   margin-top:auto;
   margin-bottom:auto;
   margin-left: auto;
   margin-right: auto;
   align:center} 

td{ padding: 6px 10px 6px 10px } 

th{ text-align: left; } 

</style>
```{r, eval=FALSE}
# Extra information
bb <- st_bbox( c( xmin = -13055498, xmax = -13011238, 
                  ymax = 3853073, ymin = 3817112 ), 
               crs = st_crs("+init=epsg:3395"))

#

dd.URL <- "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/data-dictionary.csv"
data.dictionary <- read.csv( dd.URL, stringsAsFactors=F ) 
# data.dictionary must have "LABEL" for var name and "VARIABLE" for description
data.dictionary <- data.dictionary[ -1 , ] # drop the original column names
data.label <- data.dictionary[ data.dictionary$VARIABLE %in% these.variables , ]
#
# replace these with descriptive labels 
# from the data dictionary 
temp.names <- paste0( "Variable ", these.variables )
#
# Adding interpretable variable names
# from the data dictionary:
# add a name attribute for each variable
# 
# value <- c(1,2,3)
# dd.name <- c("one","two","three")
value <- data.label$LABEL
dd.name <- data.label$VARIABLE
# x <- dd.name
# names(x) <- value
xnames <- dd.name
names(xnames) <- value
# dd names and values linked
# names( x[2] )
# can now get the label using the value
# using the name attributes 
# x[ "two" ]
#
# to add labels to the maps
# use the radio button value 
# to get the data dictionary label: 
# x[input$demographics ]
# x[input$label 
temp.names <- Short.names
#
names(button.values) <- c("Median Home Value 2000","Median Home Value 2010","Value Change 2000-2010","Growth in Home Value")
#
# Median Home Values 2000
home.value <- function(MHV2000, MHV2010, CHV0010, GHV0010)
  
home.value %>% 
  {
  MHV2000 <- 
  {
hist( mhv.00, breaks=200, xlim=c(0,500000), 
      col="gray20", border="white",
      axes=F, 
      xlab="MHV (median = $138k)",
      ylab="",
      main="Median Home Value in 2000 (2010 US dollars)" )

axis( side=1, at=seq(0,500000,100000), 
      labels=c("$0","$100k","$200k","$300k","$400k","$500k") )

abline( v=median( mhv.00, na.rm=T ), col="orange", lwd=3 ) 
renderPlot({})
}
# split the selected variable into deciles 
#
#
# Median Home Values 2010
MHV2010 <- 
  {
hist( mhv.10, breaks=200, xlim=c(0,500000), 
      col="gray20", border="white",
      axes=F, 
      xlab="MHV (median = $138k)",
      ylab="",
      main="Median Home Value in 2010 (2010 US dollars)" )

axis( side=1, at=seq(0,500000,100000), 
      labels=c("$0","$100k","$200k","$300k","$400k","$500k") )

abline( v=median( mhv.10, na.rm=T ), col="orange", lwd=3 ) 
renderPlot({})
}
# 
#
# Change in home values 2000-2010
CHV0010 <-
  {
hist( mhv.change/1000, breaks=500, 
      xlim=c(-100,500), yaxt="n", xaxt="n",
      xlab="Thousand of US Dollars (adjusted to 2010)", cex.lab=1.5,
      ylab="", main="Change in Median Home Value 2000 to 2010",
      col="gray20", border="white" )

axis( side=1, at=seq( from=-100, to=500, by=100 ), 
      labels=paste0( "$", seq( from=-100, to=500, by=100 ), "k" ) )
        
mean.x <- mean( mhv.change/1000, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=200, y=1500, 
      labels=paste0( "Mean = ", dollar( round(1000*mean.x,0)) ), 
      col="darkorange", cex=1.2, pos=3 )

median.x <- median( mhv.change/1000, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=200, y=2000, 
      labels=paste0( "Median = ", dollar( round(1000*median.x,0)) ), 
      col="dodgerblue", cex=1.2, pos=3 )
}
# 
#
# Growth in Home Value
GHV0010 <-
  {
hist( mhv.growth, breaks=5000, 
      xlim=c(-100,200), yaxt="n", xaxt="n",
      xlab="", cex.main=1.5,
      ylab="", main="Growth in Home Value by Census Tract 2000 to 2010",
      col="gray40", border="white" )

axis( side=1, at=seq( from=-100, to=200, by=50 ), 
      labels=paste0( seq( from=-100, to=200, by=50 ), "%" ) )

ymax <- max(  )
        
mean.x <- mean( mhv.growth, na.rm=T )
abline( v=mean.x, col="darkorange", lwd=2, lty=2 )
text( x=100, y=(0.5*ymax), 
      labels=paste0( "Mean = ", round(mean.x,0), "%"), 
      col="darkorange", cex=1.2, pos=4 )

median.x <- median( mhv.growth, na.rm=T )
abline( v=median.x, col="dodgerblue", lwd=2, lty=2 )
text( x=100, y=(0.6*ymax), 
      labels=paste0( "Median = ", round(median.x,0), "%"), 
      col="dodgerblue", cex=1.2, pos=4 )
  }
}

```
